---
output: html_document
number-sections: true
execute:
  echo: false
  output: false
  warning: false
editor_options: 
  chunk_output_type: console
---

This handbook uses the following packages:

```{r startup}
#| echo: true
#| output: false
#| warning: false

library(tidyverse)
library(plotly)
library(DiagrammeR)
library(readxl)
library(skimr)


```

An example dataset will be used to demonstrate the different features of R programming. You can find information about the dataset at @sec-loadfiles.

# Course content

This is the handbook for the PhD course on ***R for environmental chemistry data analysis***

<!-- Use this HTML syntax to comment out text from the rendered book. THIS TEXT WILL NOT SHOW IN THE BOOK.-->

<!-- If you use RStudio, you can use the keyboard shortcut Ctrl + Shift + C (Command + Shift + C on macOS) to comment out a line of text. -->

## **Plan for course**

The final report will be an online book with chapters with one containing a learning diary, and others should contain a scientific report documenting the descriptive analysis and visualization of environmental chemistry relevant data sets.

[Start with R software]{.underline}

<https://moderndive.com/1-getting-started.html>

-   Install R and RStudio

-   Install RTools

-Install packages

-Managing R with .Rprofile,..: <https://support.posit.co/hc/en-us/articles/360047157094-Managing-R-with-Rprofile-Renviron-Rprofile-site-Renviron-site-rsession-conf-and-repos-conf>

#### Basic R

-   Basic concepts

-   Data types: vector, dataframe, list,...

-   Using base R functions: c, paste,..

-   Using functions from installed packages

-- library()

-- Use help "?"

-   If statements, for loops

-   Basic plotting

-   Importing excel, csv,..

[Using version control with Git and Github]{.underline}

-   Version Control with Git and SVN: <https://support.posit.co/hc/en-us/articles/200532077-Version-Control-with-Git-and-SVN>

-   Happy Git and GitHub for the user: <https://happygitwithr.com/>

[Writing and creating markdown documents]{.underline}

-   Markdown syntax: <https://quarto.org/docs/authoring/markdown-basics.html>

-   Quarto Books: <https://quarto.org/docs/books/>

-   Quarto blog: <https://albert-rapp.de/posts/13_quarto_blog_writing_guide/13_quarto_blog_writing_guide.html>

#### Tidyverse

-   Tidyverse concepts

-   The pipe: %\>% or \|\>  <https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/>

-   ggplot2: <https://ggplot2.tidyverse.org/>

-   dplyr: <https://dplyr.tidyverse.org/>

-   pivot

<!-- -->

-   tidyr: <https://tidyr.tidyverse.org/>

-   readr, readxl: <https://readr.tidyverse.org/> ; <https://readxl.tidyverse.org/>

#### Interactive plotting with R

-   Javascript and htmlwidgets: <https://www.htmlwidgets.org/>

-   Plotly: <https://plotly.com/r/>  

-   DT: <https://rstudio.github.io/DT/>

#### Statistics

-   Basic stats

-   Comparison between groups (t-test, K-W, ANOVA,...)

-   Regression/correlation

-   Multivariate statistics (PCA, HCA,...)

#### ChatGPT

#### Database (optional)

-        Relational database management system, MySQL

o   <https://www.hostinger.com/tutorials/database-schema>

o   ...

-        dbplyr: https://dbplyr.tidyverse.org/

## [Recommended literature:]{.underline}

Introduction to Programming with R. <https://discdown.org/rprogramming/introduction.html>

Hands-On Programming with R. <https://rstudio-education.github.io/hopr/>

R for Data Science (2e). Hadley Wickham. Freely available at: <https://r4ds.hadley.nz/>

R for Data Science: Exercise Solutions. <https://jrnold.github.io/r4ds-exercise-solutions/>

R Graphics Cookbook, 2nd edition. Winston Chang. Freely available at: <https://r-graphics.org/>

Posit Cheatsheets: <https://posit.co/resources/cheatsheets/>

Statistical Inference via Data Science: A ModernDive into R and the Tidyverse. <https://moderndive.com/index.html>

Interactive web-based data visualization with R, plotly, and shiny. <https://plotly-r.com/>

Data Visualization: A practical introduction. Kieran Healy. <https://socviz.co/>

Mastering Software Development in R: <https://bookdown.org/rdpeng/RProgDA/>

To learn more about Quarto books visit:

-   <https://quarto.org/docs/books>

-   <https://r4ds.hadley.nz/quarto.html>

-   <https://quarto.org/docs/computations/execution-options.html>

Some useful links to other R for data science books:

-   <https://www.r4epi.com/installing-r-and-rstudio>

For advanced R users

-   <http://adv-r.had.co.nz/>

**Session information with the versions of different packages**

```{r}

sessionInfo()
```
